WSClean: an implementation of a fast, generic wide-field imager

### 2014-02-28 articles/2014/Article-WSClean/wsclean.tex
Implemented Natasha's comments

\ifpdf\pdfinfo{/Title      (WSClean)
               /Author     (A. R. Offringa et al.)
               /Keywords   (instrumentation: interferometers;methods: observational;techniques: interferometric;radio continuum: general)
        }
\else\usepackage{graphics}\fi

@@

\begin{abstract}
We present a new interferometric imager that uses the $w$-stacking algorithm and can make use of the $w$-snapshot algorithm. The performance dependencies of CASA's $w$-projection and our new imager are analysed and analytical functions are derived that describe the required computing cost for both imagers. On MWA data, we find our new method to be an order of magnitude faster than $w$-projection, as well as being capable of full-sky imaging at full resolution and with full polarization correction.  We predict the computing costs for several other arrays and estimate that our imager is a factor of 2--12 faster, depending on the array configuration. We estimate the computing cost for imaging full SKA1-low observations to be 60 PetaFLOPS with current techniques. We find that combining $w$-stacking with the $w$-snapshot algorithm does not significantly improve computing requirements. The source code of our new imager is publicly released.
\end{abstract}

@@

\section{Introduction}

Observations from non-coplanar interferometric radio telescopes that image large fractions of the sky at once can not be imaged with a two-dimensional fast Fourier transform (FFT). Instead, the imaging algorithm needs to account for the ``$w$-term'' during inversion, which is the term that describes the deviation of the array from a perfect plane \citep{perley-noncoplanar-arrays}. This deviation is amplified for telescopes with wide field of view, making this especially an issue for low-frequency telescopes that by nature are wide-field instruments.

@@

New generation of wide-field observatories are producing data sets that are orders of magnitude larger than before. Examples of such telescopes include the Murchison Widefield Array (MWA, \citealt{mwa-design-2009}, \citealt{mwa}), the upgraded Jansky Very Large Array (JVLA) and the Low-Frequency Array (LOFAR, \citealt{lofar-2013}). The Common Astronomy Software Applications (CASA, \citealt{casa-2007}, \citealt{casa-2008}) have an efficient implementation of the $w$-projection algorithm, with many available features such as multi-scale clean and spectral-shape fitting during deconvolution. However, at the MWA we have seen that imaging a 2-min snapshot observation at low-elevation can take up to tens of hours with CASA's $w$-projection algorithm. Larger images or higher zenith angles can be impossible to image because the size and number of $w$-kernels become too large to hold in memory.

@@

In this article, we present a new implementation of a generic wide-field imager that is significantly faster than CASA's $w$-projection implementation, and is able to successfully image any MWA observation. To obtain the increase in speed, the implementation uses the $w$-stacking method for correcting the $w$-terms, optionally combined with a new technique for $w$-snapshot imaging. We named the new imager ``WSClean'', as an abbreviation for ``$w$-Stacking Clean''. Our new imaging implementation, which we believe to be the fastest generic wide-field imager available, is publicly released\footnote{The WSClean source code can be found at:\\\url{http://sourceforge.net/p/wsclean}}.

@@

For small field of views, the term $\sqrt{1-l^2-m^2}$ is approximately of unit size, making Eq.~\eqref{eq:visibility-function} approximate an ordinary 2-dimensional Fourier transform that can be inverted with an inverse FFT. A common rule is that this is valid when
\begin{equation}\label{eq:when-2d-is-valid}
\forall w,l,m: w\left(\sqrt{1-l^2-m^2}-1\right) \ll 1.
\end{equation}

To derive the $w$-stacking technique, Eq.~\eqref{eq:visibility-function} is rewritten to
\begin{align}\notag
V(u,v,w) = & \iint \frac{I'(l,m) e^{-2\pi i w(\sqrt{1-l^2-m^2}-1)}}{\sqrt{1-l^2-m^2}} \times \\ \notag
& e^{-2\pi i \left(ul + vm\right)} dl dm.
\end{align}
This is an ordinary two-dimensional Fourier transform going from $u,v$ space to $l,m$ space, and can be inverted to get:
% Intermediate step:
%\begin{align}\notag
%\frac{I'(l,m) e^{-2\pi i w(\sqrt{1-l^2-m^2}-1)}}{\sqrt{1-l^2-m^2}} = & \iint V(u,v,w) \times \\ \notag
%& e^{-2\pi i \left(ul + vm\right)} du dv,
%\end{align}
%which can be rewritten to
\begin{align}\notag
\frac{I'(l,m)}{\sqrt{1-l^2-m^2}} = & e^{2\pi i w(\sqrt{1-l^2-m^2}-1)} \iint V(u,v,w) \times \\ \notag
& e^{2\pi i \left(ul + vm\right)} du dv.
\end{align}
Integrating both sides over $w_{\min}$ to $w_{\max}$, the minimum and maximum value of $w$, results in
\begin{align}\notag
\frac{I'(l,m)\left(w_{\max} - w_{\min}\right)}{\sqrt{1-l^2-m^2}} = \int\limits_{w_{\min}}^{w_{\max}} e^{2\pi i w(\sqrt{1-l^2-m^2}-1)} \times \\ \label{eq:wstacking}
\iint V(u,v,w)  e^{2\pi i \left(ul + vm\right)} du dv dw.
\end{align}
The final step is to make the $u,v,w$ parameters discrete, so that the integration over $u$ and $v$ can become an inverse FFT and the integration over $w$ becomes a summation. This shows that the sky function can be reconstructed by i)~gridding samples with equal $w$-value on a uniform grid; ii)~calculating the inverse FFT; iii)~apply the direction-dependent phase shift $e^{2\pi i w(\sqrt{1-l^2-m^2}-1)}$; iv)~repeat this for all $w$-values and add the results together; v)~apply the final scaling.

In practice, the final scaling will be different from $\left(w_{\max} - w_{\min}\right)/\sqrt{1-l^2-m^2}$ suggested by Eq.~\eqref{eq:wstacking}, because the individual $w$-layers will not be completely filled with samples. Therefore, the number of added samples and their weight have to be taken into account. Additionally, it might be required to divide out the effect of a possible convolution kernel, the primary beam and correct for other direction-dependent effects. In XX, YY, LL or RR polarizations, a correlated baseline is the complex conjugate of the reversed baseline, and the relation $V(u,v,w)=\overline{V(-u,-v,-w)}$ holds. The right-hand side of Eq.\eqref{eq:wstacking} with only positive $w$-value samples then becomes the complex conjugate of the one with only negative $w$-values samples. In this case we can therefore calculate the image for $w<0$ from the image with $w>0$. That allows us to set $w_{\min}$ to the minimum absolute $w$-value, which requires two times fewer layers. In any case, the input to the two-dimensional inverse FFT is not generally an Hermitian-symmetric function, and hence the inverse FFT is always performed as a complex-to-complex transform.

@@

\subsection{Discretization of $w$} \label{sec:gridding-w}
\begin{figure}
\begin{center}
\includegraphics[width=8cm]{img/aliasing-example-annotated}
\caption{Aliasing artefacts caused by insufficient $w$-layers in a simulated field. WSClean was set to use 12 $w$-layers. The centre of the image is at 10\degree zenith angle, which would normally require $\sim195$ $w$-layers. Sources are 1~Jy (red circles), ghost sources are approximately 0.2~Jy. Each source produces two ghost sources, but because they re-appear after a major cleaning cycle, they are eventually cleaned and produce more ghost sources. }
\label{fig:aliasing-example}
\end{center}
\end{figure}
While the discretization of $u$ and $v$ is similar to conventional imaging, the discretization of $w$ defines the number of $w$-layers that need to be processed. For this, one can use a rule similar to \eqref{eq:when-2d-is-valid}, and make sure that the phase difference for two subsequent discretized $w$-values, $w_A$ and $w_B$,  is less than one radian. This results in the constraint
\begin{equation} \label{eq:minimum-w-distance}
\left|\left(w_A - w_B\right) 2\pi (\sqrt{1-l^2-m^2}-1)\right| \ll 1.
\end{equation}
This suggest that a uniform discretization in $w$ is optimal. This is in contrast to \citet{wprojection-cornwell} where $\sqrt{w}$ tabulation is suggested. From Eq.~\eqref{eq:minimum-w-distance}, the required number of layers can be derived and is given by
\begin{equation} \label{eq:nwlayers-bound}
 N_\textrm{wlayers} \gg 2\pi \left(w_{\max} - w_{\min}\right) \max_{l,m} \left(1 - \sqrt{1-l^2-m^2}\right).
\end{equation}
Actual values for the right-hand side can differ a lot. The value of $w_{\max} - w_{\min}$ is influenced by the coplanarity of the array, the zenith angle and the wavelength, while the value of the $\max_{l,m}$ term is influenced by the angular size of the image. For the MWA, a typical value of $w_{\max} - w_{\min}$ is $\sim 10$ at zenith, but is already $\sim 400$ at a zenith angle of $30^{\circ}$. For a typical full-field-of-view image of a MWA observation of $3072\times 3072$ pixels of $0.75''$ size, the $\max_{l,m}$ term is $0.68$. This implies that tens of $w$-layers are required at zenith and hundreds at lower elevations. The number of $w$-layers has a large effect on the performance of the $w$-stacking algorithm, and will be discussed further in the next sections.

@@

\subsection{Time complexity of $w$-stacking} \label{sec:time-complexity}
It is useful to analyse the time complexity of $w$-stacking and compare it with $w$-projection, to understand which algorithm performs better in a given situation. We will use the following symbols: $N_\textrm{wlay}$ is the number of $w$-layers for $w$-stacking, $N_\textrm{pix}$ is the number of pixels in the image along each side, $N_\textrm{vis}$ is the number of visibilities, $N_\textrm{kern}$ is the size of the anti-aliasing kernel (see \S\ref{sec:gridding}), $N_\textrm{wkern}$ is the size of the $w$-kernel for $w$-projection, $w_{\max}$ is the maximum $w$ value and $\alpha_\textrm{FOV}$ is the imaging field of view.

@@

Table~\ref{tbl:computational-cost-per-operation} shows how the computational costs scale for the operations that dominate the imaging in the $w$-stacking and $w$-projection algorithms. For comparison, we can assume the gridding kernel adds a constant term and the terms $N_\textrm{wlay}$ and $N_\textrm{wkern}$ follow approximately $w_{\max} \sin \alpha_\textrm{FOV}$. The time complexity for $w$-stacking is then given by
\begin{equation}
\textrm{TC}_\textrm{wstacking}=\mathcal{O}\left(N^2_\textrm{pix} \log N_\textrm{pix} w_{\max} \sin \alpha_\textrm{FOV}+ N_\textrm{vis} \right),
\end{equation}
and for $w$-projection it is
\begin{equation}
\textrm{TC}_\textrm{wprojection}=\mathcal{O}\left(N^2_\textrm{pix} \log N_\textrm{pix} + N_\textrm{vis} w_{\max}^2 \sin^2 \alpha_\textrm{FOV}\right).
\end{equation}
From these bounds it can be concluded that in the limiting behaviour, the $w$-stacking method will be faster when the gridding of the visibilities is the dominating cost of the algorithm. The $w$-projection algorithm will be faster when the inverse FFTs are dominating. In Sect.~\ref{sec:analysis} we will determine which method is faster in practice for different parameters.

\section{The WSClean imager implementation} \label{sec:implementation}
With the purpose of testing new algorithms for the imaging of MWA data, we have written a new imager around the $w$-stacking algorithm. The imager is not specialized for the MWA, and has been successfully used for imaging data from other telescopes.

The new imager, called ``WSClean'', is written in the C++ language. It reads visibilities from CASA measurement sets and writes output images to Flexible Image Transport System (FITS) files. Several steps are multi-threaded using the threading module of the C++11 standard library. These are: reading and writing; gridding from and to different $w$-layers; performing the FFTs; and H\"ogbom Clean iterations (peak-finding and image subtraction, \citealt{hogbom-clean}). For the latter, intrinsics are used as well. Because of these optimizations, minor Cleaning iterations with an image size of $3072\times3072$ are performed at a rate of hundreds per second. This is fast enough to make the Clark Clean optimization \citep{clark-clean}, which consists of considering only a subset of pixels with a reduced PSF, less relevant. For very large images this optimization might still be useful, but we have not implemented it in WSClean. Because the PSF varies for imaging with non-zero $w$-values, subtracting a constant PSF in image space leads to inaccuracies. After a number of minor iterations, it is therefore beneficial to invert the model back to the visibilities via prediction, and subtract the model directly from the visibilities \citep{wprojection-cornwell}. This is similar to the Cotton-Schwab cleaning method \citep{cotton-schwab-clean}. WSClean allows cleaning individual polarizations, or can jointly deconvolve the polarizations. In the latter case, peaks are searched in $pp^2+2pq \overline{(pq)} + qq^2$ space, where $\overline{pq}=qp$ is the complex conjugate of $pq$, and the PSF is subtracted from the polarizations with different factors.

It will not be possible to store all $w$-layers in memory when creating large images or when many $w$-layers are used. For example, our 32-GB test machine can store 227 $w$-layers of $3072\times3072$ in memory. In more demanding imaging configurations, the implementation performs several passes over the measurement set and will grid a subset of $w$-layers in each pass.

@@

To efficiently image the full sky, an observation is split into short snapshots, their phase centre is changed to zenith and subsequently a suitable area is imaged. Depending on desired resolution and dynamic range, this might require very large images. To make an image at the MWA resolution, the image needs to have approximately 10k pixels along each side. The $w$-values will be very small at zenith, because at zenith only the vertical offsets of the antennas will contribute to the $w$-value. For the 128-tile MWA, the maximum differential elevation between tiles is $8.5$m. The tiles are in fact on a slight slope, and by fitting a plane to the antennas and changing the phase centre towards the normal of that plane, the maximum $w$-value decreases to $5.5\textrm{m} / \lambda$. In the following sections, when discussing the MWA zenith direction we are referring to this optimal $w$-direction.

\subsection{$w$-snapshots} \label{sec:snapshot-imaging}
\begin{figure*}
\begin{center}
\includegraphics[width=8cm]{img/vela-normal-projection}
\includegraphics[width=8cm]{img/vela-zenith-projection}
\caption{13-min MWA observation of supernova remnants Vela and Puppis A. Left: normal projection centred on Puppis A, right: phase-rotated to zenith to reduce $w$-values, recentred on Puppis A during imaging. Both Stokes-I images have been made with WSClean using the Cotton-Schwab clean algorithm. No beam correction was applied. Imaging computing cost was 60 min for the normal projected image and 41 min for the recentred image.}
\label{fig:vela-projection-example}
\end{center}
\end{figure*}
As discussed, all-sky snapshot imaging with zenith as phase centre is quite efficient. However, if one is not interested in imaging the whole sky, and the direction of interest is far from zenith, the computational overhead of making all-sky images is undesirable.

Instead of imaging the whole sky, the image can be recentred from $(l,m)$ to $(\hat{l},\hat{m})=(l+\Delta l,m+\Delta m)$ by performing the substitution $(l,m)\leftarrow(\hat{l},\hat{m})$ in Eq.~\eqref{eq:wstacking}:
\begin{align}\notag
\frac{I'(\hat{l}, \hat{m})\left(w_{\max} - w_{\min}\right)}{\sqrt{1-\hat{l}^2-\hat{m}^2}} = \int\limits_{w_{\min}}^{w_{\max}} e^{2\pi i w(\sqrt{1-\hat{l}^2-\hat{m}^2}-1)} \times \\
\iint e^{2 \pi i \left(u\Delta l + v\Delta m\right)} V(u,v,w) e^{2\pi i \left(ul + vm\right)} du dv dw.
\end{align}
In words, recentering an image involves accounting for the position shift during the $w$-correction and final scaling, and shifting the visibilities in phase by multiplication with $e^{2 \pi i \left(u\Delta l + v\Delta m\right)}$ prior to the imaging. This essentially is a method of implementing the $w$-snapshots algorithm as described by \citet{widefield-imaging-ska-cornwell}. However, instead of creating warped images by performing the FFT over a tilted plane in $uvw$-space, our method transforms planes with constant $w$-values. The benefit of our method is that the resulting image has a circular synthesized beam, i.e., the resolution in $l$ and $m$ directions matches the intrinsic resolution of the instrument, which is desirable for cleaning. Our method is also relatively easy to implement. It has a similar computational cost compared to the $w$-snapshot algorithm, although the required number of $w$-layers has a different dependency on zenith angle, field of view and the non-coplanarity.

@@

A recentred image is in the projection of the zenith tangent plane, and will need an additional regridding step before multiple snapshots can be added together. A recentred image can be stored in a Flexible Image Transport System (FITS) file with the slant orthographic (SIN) projection normally used in interferometric imaging \citep{fits-coordinates-2002}, by setting the centre of tangent projection with the \texttt{CRPIXi} keyword \citep{wcs-in-fits}. Common viewers such as Kvis and DS9 support such FITS files and display their coordinates correctly. Unlike warped snapshots, recentred images do not need the generalized-SIN-projection keywords \texttt{PV2\_1} and \texttt{PV2\_2}.

An example of the difference in projection between non-recentred and recentred imaging is given in Fig.~\ref{fig:vela-projection-example}, which displays a 13-min MWA observation imaged with WSClean. The processing steps performed for this image were i) preprocessing of the observation with the Cotter preprocessing pipeline, which include time averaging and RFI flagging with the AOFlagger \citep{post-correlation-rfi-classification,scale-invariant-rank-operator}; ii) calibrate on Hydra~A without direction dependence using a custom implementation of the RTS full-polarization calibration algorithm \citep{rts-mwa}; and iii) image each of the seven 112 s intervals separately. Cleaning an image that is in a different projection yields slightly different results, but qualitatively the two images are clearly of equal accuracy. The wall-clock time for imaging is 60 min and 41 min for respectively normal projection and the recentred image. Of the 41 min, 3 min is spent on phase rotating the visibilities.

Another method to shift an image is by using the periodicity of the FFT function. Since sources outside the field of view will be aliased back into the field, the image can be transformed to have the alias in the centre. This complicates gridding during imaging, because the anti-aliasing kernel needs to have a pass-band shape instead of a low-pass shape. Therefore, we chose to implement the former method.

>>>>

\ifpdf\pdfinfo{/Title      (WSClean: an implementation of a fast, generic wide-field imager)
               /Author     (A. R. Offringa et al.)
               /Keywords   (instrumentation: interferometers;methods: observational;techniques: interferometric;radio continuum: general)
        }
\else\usepackage{graphics}\fi

@@

\begin{abstract}
We present a new interferometric imager that uses the $w$-stacking algorithm and can make use of the $w$-snapshot algorithm. The performance dependencies of \textsc{CASA}'s $w$-projection and our new imager are analysed and analytical functions are derived that describe the required computing cost for both imagers. On data from the Murchison Widefield Array, we find our new method to be an order of magnitude faster than $w$-projection, as well as being capable of full-sky imaging at full resolution and with correct polarization correction.  We predict the computing costs for several other arrays and estimate that our imager is a factor of 2--12 faster, depending on the array configuration. We estimate the computing cost for imaging the low-frequency Square-Kilometre Array observations to be 60 PetaFLOPS with current techniques. We find that combining $w$-stacking with the $w$-snapshot algorithm does not significantly improve computing requirements. The source code of our new imager is publicly released.
\end{abstract}

@@

\section{Introduction}
Visibility data from non-coplanar interferometric radio telescopes which observe large fractions of the sky at once can not be accurately imaged with a two-dimensional fast Fourier transform (FFT). Instead, the imaging algorithm needs to account for the ``$w$-term'' during inversion, which is the term that describes the deviation of the array from a perfect plane \citep{perley-noncoplanar-arrays}. This deviation is amplified for telescopes with wide field of view (FOV), making this a particularly important issue for low-frequency telescopes that by nature are wide-field instruments.

@@

New generation of wide-field observatories are producing data sets that are orders of magnitude larger than before. Examples of such telescopes include the Murchison Widefield Array (MWA, \citealt{mwa-design-2009}, \citealt{mwa}), the upgraded Jansky Very Large Array (JVLA) and the Low-Frequency Array (LOFAR, \citealt{lofar-2013}). The Common Astronomy Software Applications (\textsc{CASA}, \citealt{casa-2007}, \citealt{casa-2008}) have an efficient implementation of the $w$-projection algorithm, with many available features such as multi-scale clean and spectral-shape fitting during deconvolution. However, with the MWA we have seen that imaging a 2-min snapshot observation at low-elevation can take up to tens of hours with \textsc{CASA}'s $w$-projection algorithm. Larger images or higher zenith angles can be impossible to image because the size and number of $w$-kernels become too large to hold in memory.

@@

In this article, we present a new implementation of a generic wide-field imager that is significantly faster than \textsc{CASA}'s $w$-projection implementation, and is able to successfully image any MWA observation. To obtain the increase in speed, the implementation uses the $w$-stacking method for correcting the $w$-terms, optionally combined with a new technique for $w$-snapshot imaging. We named the new imager ``\textsc{WSClean}'', as an abbreviation for ``$w$-Stacking Clean''. Our new imaging implementation, which we believe to be the fastest generic wide-field imager available, is publicly released\footnote{The \textsc{WSClean} source code can be found at:\\\href{http://sourceforge.net/p/wsclean}{http://sourceforge.net/p/wsclean}}.

@@

For small FOVs, the term $\sqrt{1-l^2-m^2}$ is approximately of unit size, making Eq.~\eqref{eq:visibility-function} approximately an ordinary invertable two-dimensional Fourier transform. A common rule is that this is valid when
\begin{equation}\label{eq:when-2d-is-valid}
\forall w,l,m: w\left(\sqrt{1-l^2-m^2}-1\right) \ll 1.
\end{equation}

To derive the $w$-stacking technique, Eq.~\eqref{eq:visibility-function} is rewritten to
\begin{align}\notag
V(u,v,w) = & \iint \frac{I'(l,m) e^{-2\pi i w(\sqrt{1-l^2-m^2}-1)}}{\sqrt{1-l^2-m^2}} \times \\ \notag
& e^{-2\pi i \left(ul + vm\right)} dl dm.
\end{align}
This is an ordinary two-dimensional Fourier transform going from $u,v$ space to $l,m$ space, and can be inverted to get:
% Intermediate step:
%\begin{align}\notag
%\frac{I'(l,m) e^{-2\pi i w(\sqrt{1-l^2-m^2}-1)}}{\sqrt{1-l^2-m^2}} = & \iint V(u,v,w) \times \\ \notag
%& e^{-2\pi i \left(ul + vm\right)} du dv,
%\end{align}
%which can be rewritten to
\begin{align}\notag
\frac{I'(l,m)}{\sqrt{1-l^2-m^2}} = & e^{2\pi i w(\sqrt{1-l^2-m^2}-1)} \iint V(u,v,w) \times \\ \notag
& e^{2\pi i \left(ul + vm\right)} du dv.
\end{align}
Integrating both sides over $w_{\min}$ to $w_{\max}$, the minimum and maximum value of $w$, results in
\begin{align}\notag
\frac{I'(l,m)\left(w_{\max} - w_{\min}\right)}{\sqrt{1-l^2-m^2}} = \int\limits_{w_{\min}}^{w_{\max}} e^{2\pi i w(\sqrt{1-l^2-m^2}-1)} \times \\ \label{eq:wstacking}
\iint V(u,v,w)  e^{2\pi i \left(ul + vm\right)} du dv dw.
\end{align}
The final step is to make the $u,v,w$ parameters discrete, so that the integration over $u$ and $v$ can become an inverse FFT and the integration over $w$ becomes a summation. This shows that the sky function can be reconstructed by i)~gridding samples with equal $w$-value on a uniform grid; ii)~calculating the inverse FFT; iii)~applying the direction-dependent phase shift $e^{2\pi i w(\sqrt{1-l^2-m^2}-1)}$; iv)~repeating this for all $w$-values and adding the results together; v)~applying the final scaling.

In practice, the final scaling will be different from $\left(w_{\max} - w_{\min}\right)/\sqrt{1-l^2-m^2}$ suggested by Eq.~\eqref{eq:wstacking}, because the individual $w$-layers will not be completely filled with samples. Therefore, the number of added samples and their weight have to be taken into account. Additionally, it might be required to divide out the effect of a possible convolution kernel, the primary beam and correct for other direction-dependent effects. In XX, YY, LL or RR polarizations, a correlated baseline is the complex conjugate of the reversed baseline, and the relation $V(u,v,w)=\overline{V(-u,-v,-w)}$ holds. The right-hand side of Eq.\eqref{eq:wstacking} with only positive $w$-value samples then becomes the complex conjugate of the one with only negative $w$-value samples. In this case we can therefore calculate the image for $w<0$ from the image with $w>0$. That allows us to set $w_{\min}$ to the minimum absolute $w$-value, which requires a factor of two fewer layers. In any case, the input to the two-dimensional inverse FFT is not generally an Hermitian-symmetric function, and hence the inverse FFT is always performed as a complex-to-complex transform.

@@

\subsection{Discretization of $w$} \label{sec:gridding-w}
\begin{figure}
\begin{center}
\includegraphics[width=8cm]{img/aliasing-example-annotated}
\caption{Aliasing artefacts caused by insufficient $w$-layers in a simulated field. \textsc{WSClean} was set to use 12 $w$-layers. The centre of the image is at 10\degree zenith angle, which would normally require $\sim195$ $w$-layers. Sources are 1~Jy (red circles), ghost sources are approximately 0.2~Jy. Each source produces two ghost sources, but because they re-appear after a major cleaning cycle, they are eventually cleaned and produce more ghost sources. }
\label{fig:aliasing-example}
\end{center}
\end{figure}
While the discretization of $u$ and $v$ is similar to conventional imaging, the discretization of $w$ defines the number of $w$-layers that need to be processed. For this, one can use a rule similar to \eqref{eq:when-2d-is-valid}, and make sure that the phase difference for two subsequent discretized $w$-values, $w_A$ and $w_B$,  is less than one radian. This results in the constraint
\begin{equation} \label{eq:minimum-w-distance}
\left|\left(w_A - w_B\right) 2\pi (\sqrt{1-l^2-m^2}-1)\right| \ll 1.
\end{equation}
This suggest that a uniform discretization in $w$ is optimal. This is in contrast to \citet{wprojection-cornwell} where $\sqrt{w}$ tabulation is suggested. From Eq.~\eqref{eq:minimum-w-distance}, the required number of layers can be derived and is given by
\begin{equation} \label{eq:nwlayers-bound}
 N_\textrm{wlay} \gg 2\pi \left(w_{\max} - w_{\min}\right) \max_{l,m} \left(1 - \sqrt{1-l^2-m^2}\right).
\end{equation}
Actual values for the right-hand side can be very different, depending on the observation. The value of $w_{\max} - w_{\min}$ is influenced by the coplanarity of the array, the zenith angle and the wavelength, while the value of the $\max_{l,m}$ term is influenced by the angular size of the image. For the MWA, a typical value of $w_{\max} - w_{\min}$ is $\sim 10$ at zenith, and reaches $\sim 400$ at a zenith angle of $30^{\circ}$. For a typical full-field-of-view image of a MWA observation of $3072\times 3072$ pixels of $0.75''$ size, the $\max_{l,m}$ term is $0.68$. This implies that tens of $w$-layers are required at zenith and hundreds at lower elevations. The number of $w$-layers has a large effect on the performance of the $w$-stacking algorithm, and will be discussed further in the next sections.

@@

\subsection{Time complexity of $w$-stacking} \label{sec:time-complexity}
It is useful to analyse the time complexity of $w$-stacking and compare it with $w$-projection, to understand which algorithm performs better in a given situation. We will use the following symbols: $N_\textrm{wlay}$ is the number of $w$-layers for $w$-stacking, $N_\textrm{pix}$ is the number of pixels in the image along each side, $N_\textrm{vis}$ is the number of visibilities, $N_\textrm{kern}$ is the size of the anti-aliasing kernel (see \S\ref{sec:gridding}), $N_\textrm{wkern}$ is the size of the $w$-kernel for $w$-projection, $w_{\max}$ is the maximum $w$ value and $\alpha_\textrm{FOV}$ is the imaging FOV.

@@

Table~\ref{tbl:computational-cost-per-operation} shows how the computational costs scale for the operations that dominate the imaging in the $w$-stacking and $w$-projection algorithms. For comparison, we can assume the gridding kernel adds a constant term and the terms $N_\textrm{wlay}$ and $N_\textrm{wkern}$ follow approximately $w_{\max} \sin \alpha_\textrm{FOV}$. The time complexity for $w$-stacking is then given by
\begin{equation}
\textrm{TC}_\textrm{wstacking}=\mathcal{O}\left(N^2_\textrm{pix} \log N_\textrm{pix} w_{\max} \sin \alpha_\textrm{FOV}+ N_\textrm{vis} \right),
\end{equation}
and for $w$-projection it is
\begin{equation}
\textrm{TC}_\textrm{wprojection}=\mathcal{O}\left(N^2_\textrm{pix} \log N_\textrm{pix} + N_\textrm{vis} w_{\max}^2 \sin^2 \alpha_\textrm{FOV}\right).
\end{equation}
From these bounds it can be concluded that in the limiting behaviour, the $w$-stacking method will be faster when the gridding of the visibilities is the dominating cost of the algorithm. The $w$-projection algorithm will be faster when the inverse FFTs are the dominant expense. In Sect.~\ref{sec:analysis} we will determine which method is faster in practice for different parameters.

\section{The \textsc{WSClean} imager implementation} \label{sec:implementation}
With the purpose of testing new algorithms for the imaging of MWA data, we have written a new imager around the $w$-stacking algorithm. The imager is not specialized for the MWA, and has been successfully used for imaging data from other telescopes.

The new imager, called ``\textsc{WSClean}'', is written in the C++ language. It reads visibilities from \textsc{CASA} measurement sets and writes output images to Flexible Image Transport System (FITS) files. Several steps are multi-threaded using the threading module of the C++11 standard library. These are: reading and writing; gridding from and to different $w$-layers; performing the FFTs; and H\"ogbom Clean iterations (peak-finding and image subtraction, \citealt{hogbom-clean}). For the latter, intrinsics are used as well. Because of these optimizations, minor Cleaning iterations with an image size of $3072\times3072$ are performed at a rate of hundreds per second. This is fast enough to make the Clark Clean optimization \citep{clark-clean}, which consists of considering only a subset of pixels with a reduced PSF, less relevant. For very large images this optimization might still be useful, but we have not implemented it in \textsc{WSClean}. Because the PSF varies for imaging with non-zero $w$-values, subtracting a constant PSF in image space leads to inaccuracies. After a number of minor iterations, it is therefore beneficial to invert the model back to the visibilities via prediction, and subtract the model directly from the visibilities \citep{wprojection-cornwell}. This is similar to the Cotton-Schwab cleaning method \citep{cotton-schwab-clean}. \textsc{WSClean} allows cleaning individual polarizations, or can jointly deconvolve the polarizations. In the latter case, peaks are searched in $pp^2+2pq \overline{(pq)} + qq^2$ space, where $\overline{pq}=qp$ is the complex conjugate of $pq$, and the PSF is subtracted from the polarizations with different factors.

It will not be possible to store all $w$-layers in memory when creating large images or when many $w$-layers are used. For example, our test machine, with 32 GB of memory, can store 227 $w$-layers of $3072\times3072$. In more demanding imaging configurations, the implementation performs several passes over the measurement set and will grid a subset of $w$-layers in each pass.

@@

To efficiently image the full sky, an observation is split into short snapshots, each of which is phase-rotated to zenith and subsequently imaged, with an appropriate number of pixels and resolution. Depending on desired resolution and dynamic range, this might require very large images. To make an image at the MWA resolution, the image needs to have approximately 10k pixels along each side. The $w$-values will be very small at zenith, because at zenith only the vertical offsets of the antennas will contribute to the $w$-value. For the 128-tile MWA, the maximum differential elevation between tiles is $8.5$m. The tiles are in fact on a slight slope, and by fitting a plane to the antennas and changing the phase centre towards the normal of that plane, the maximum $w$-value decreases to $5.5\textrm{m} / \lambda$. In the following sections, when discussing the MWA zenith direction we are referring to this optimal $w$-direction.

\subsection{$w$-snapshots} \label{sec:snapshot-imaging}
\begin{figure*}
\begin{center}
\includegraphics[width=8cm]{img/vela-normal-projection}
\includegraphics[width=8cm]{img/vela-zenith-projection}
\caption{13-min MWA observation of supernova remnants Vela and Puppis A. Left: normal projection centred on Puppis A, right: phase-rotated to zenith to reduce $w$-values, recentred on Puppis A during imaging. Both Stokes-I images have been made with \textsc{WSClean} using the Cotton-Schwab clean algorithm. No beam correction was applied. Imaging computing cost was 60 min for the normal projected image and 41 min for the recentred image.}
\label{fig:vela-projection-example}
\end{center}
\end{figure*}
As discussed, all-sky snapshot imaging with zenith as phase centre is quite efficient. However, if one is not interested in imaging the whole sky, and the direction of interest is far from zenith, the computational overhead of making all-sky images is undesirable.

Instead of imaging the whole sky, the image can be recentred from $(l,m)$ to $(\hat{l},\hat{m})=(l+\Delta l,m+\Delta m)$ by performing the substitution $(l,m)\leftarrow(\hat{l},\hat{m})$ in Eq.~\eqref{eq:wstacking}:
\begin{align}\notag
\frac{I'(\hat{l}, \hat{m})\left(w_{\max} - w_{\min}\right)}{\sqrt{1-\hat{l}^2-\hat{m}^2}} = \int\limits_{w_{\min}}^{w_{\max}} e^{2\pi i w(\sqrt{1-\hat{l}^2-\hat{m}^2}-1)} \times \\
\iint e^{2 \pi i \left(u\Delta l + v\Delta m\right)} V(u,v,w) e^{2\pi i \left(ul + vm\right)} du dv dw.
\end{align}
In words, recentering an image involves accounting for the position shift during the $w$-correction and final scaling, and shifting the visibilities in phase by multiplication with $e^{2 \pi i \left(u\Delta l + v\Delta m\right)}$ prior to the imaging. This essentially is a method of implementing the $w$-snapshots algorithm as described by \citet{widefield-imaging-ska-cornwell}. However, instead of creating warped images by performing the FFT over a tilted plane in $uvw$-space, our method transforms planes with constant $w$-values. The benefit of our method is that the resulting image has a circular synthesized beam, i.e., the resolution in $l$ and $m$ directions matches the intrinsic resolution of the instrument, which is desirable for cleaning. Our method is also relatively easy to implement. It has a similar computational cost compared to the $w$-snapshot algorithm, although the required number of $w$-layers has a different dependency on zenith angle, FOV and the non-coplanarity.

@@

A recentred image is in the projection of the zenith tangent plane, and will need an additional regridding step before multiple snapshots can be added together. A recentred image can be stored in a Flexible Image Transport System (FITS) file with the slant orthographic (SIN) projection normally used in interferometric imaging \citep{fits-coordinates-2002}, by setting the centre of tangent projection with the \texttt{CRPIXi} keyword \citep{wcs-in-fits}. Common viewers such as \textsc{Kvis} and \textsc{DS9} support such FITS files and display their coordinates correctly. Unlike warped snapshots, recentred images do not need the generalized-SIN-projection keywords \texttt{PV2\_1} and \texttt{PV2\_2}.

An example of the difference in projection between non-recentred and recentred imaging is given in Fig.~\ref{fig:vela-projection-example}, which displays a 13-min MWA observation imaged with \textsc{WSClean}. The processing steps performed for this image were i) preprocessing of the observation with the Cotter preprocessing pipeline, which include time averaging and RFI flagging with the AOFlagger \citep{post-correlation-rfi-classification,scale-invariant-rank-operator}; ii) calibration using Hydra~A without direction dependence using a custom implementation of the RTS full-polarization calibration algorithm \citep{rts-mwa}; and iii) imaging of the seven 112 s intervals separately. Cleaning an image that is in a different projection yields slightly different results, but qualitatively the two images are clearly of equal accuracy. The wall-clock time for imaging is 60 min and 41 min for respectively normal projection and the recentred image. Of the 41 min, 3 min is spent on phase rotating the visibilities.

Another method to shift an image is by using the periodicity of the FFT function. Since sources outside the FOV will be aliased back into the field, the image can be transformed to have the alias in the centre. This complicates gridding during imaging, because the anti-aliasing kernel needs to have a pass-band shape instead of a low-pass shape. Therefore, we chose to implement the former method.
